% Module 1 section file. Included by ../module1_stellarator_equilibrium_geometry.tex.
\renewcommand{\refname}{References and further reading}
\phantomsection
\addcontentsline{toc}{section}{References and further reading}
\label{sec:references}
\begin{thebibliography}{99}

\bibitem{freidberg2014}
J. P. Freidberg,
\textit{Ideal MHD},
Cambridge University Press, Cambridge, 2014.

\bibitem{dHaeseleer1991}
W. D. D'haeseleer, W. N. G. Hitchon, J. D. Callen, and J. L. Shohet,
\textit{Flux Coordinates and Magnetic Field Structure: A Guide to a Fundamental Tool of Plasma Theory},
Springer, Berlin, 1991.

\bibitem{helander2014}
P. Helander,
``Theory of plasma confinement in non-axisymmetric magnetic fields,''
\textit{Reports on Progress in Physics}, vol. 77, 087001, 2014.

\bibitem{helanderSigmar2002}
P. Helander and D. J. Sigmar,
\textit{Collisional Transport in Magnetized Plasmas},
Cambridge University Press, Cambridge, 2002.

\bibitem{wesson2011}
J. Wesson,
\textit{Tokamaks}, 4th ed.,
Oxford University Press, Oxford, 2011.

\bibitem{boozer1981}
A. H. Boozer,
``Plasma equilibrium with rational magnetic surfaces,''
\textit{Physics of Fluids}, vol. 24, pp. 1999--2003, 1981.

\bibitem{hirshmanWhitson1983}
S. P. Hirshman and J. C. Whitson,
``Steepest-descent moment method for three-dimensional magnetohydrodynamic equilibria,''
\textit{Physics of Fluids}, vol. 26, pp. 3553--3568, 1983.

\bibitem{hudson2012}
S. R. Hudson, R. L. Dewar, G. Dennis, M. J. Hole, M. McGann, G. von Nessi, and S. Lazerson,
``Computation of multi-region relaxed magnetohydrodynamic equilibria,''
\textit{Physics of Plasmas}, vol. 19, 112502, 2012.

\bibitem{landremanPaul2022}
M. Landreman and E. Paul,
``Magnetic fields with precise quasisymmetry for plasma confinement,''
\textit{Physical Review Letters}, vol. 128, 035001, 2022.

\bibitem{hazeltineMeiss2003}
R. D. Hazeltine and J. D. Meiss,
\textit{Plasma Confinement},
Dover Publications, Mineola, New York, 2003.

\end{thebibliography}

